---
title: "Age Comparisons in FSA"
author: "Derek H. Ogle"
date: "`r Sys.Date()`"
format: 
  html: 
    toc: true
    toc-depth: 2
    reference-location: margin
    fig-cap-location: bottom
    tbl-cap-location: top
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    fig.width: 4
    fig.height: 4

---

# Introduction
Information about the age of fish is used to determine rates of mortality, growth, and maturation, which are key components of population models used to make fisheries management decisions. Ages of fish are often estimated by assessing patterns found in calcified structures such as scales, fin rays or spines, and otoliths. Ages estimated from calcified structures may not equal the true age of the fish, may not be the same among different technicians, or may not be the same between structures. Given the importance of age data in fisheries management, it is critical to understand sources of variability, precision, and bias in age estimates.^[These are common terms that are discussed elsewhere, including in [Ogle (2016)](https://derekogle.com/IFAR/).]

The following packages are used herein.

```{r}
#| label: setup
#| results: hide
#| message: false
library(FSA)
library(FSAdata) # for ShadCR data
library(ggplot2) # for graphing
```

In addition, the theme for plots made with `ggplot2` is set to `theme_bw()` for all plots. Further modifications to the them can be made but are not here to keep the code as simple as possible.^[For example, see [this post](https://fishr-core-team.github.io/fishR/blog/posts/2022-12-22_AFS_Style_Figures/).]

```{r}
set_theme(theme_bw())
```


# Data Requirements
The true age of each fish must be known to determine the accuracy of age estimates. Bias among structures or technicians is determined from paired age estimates for the same fish. Finally, precision among structures or technicians is examined with two *or more* age estimates for the same fish.

The data for the example used in [Ogle (2016)](https://derekogle.com/IFAR/) is in `ShadCR` distributed with `FSAdata`.^[More information about the date [is here](https://fishr-core-team.github.io/FSAdata/reference/ShadCR.html).] The true ages (`trueAge`) for fish in this sample were known because the Shad had been marked prior to being stocked. Additionally, three biologists twice (independently) estimated the age from scales for each fish. The estimated age variables are labeled with `ager`, a letter for the three biologists (`A`, `B`, or `C`) and a number for which time the scale was interpreted (`1` or `2`). Some biologists chose not to assign an age to some scales and, thus, those data are missing (shown as `NA` values).

```{r}
data("ShadCR",package="FSAdata")  # retrieve data
peek(ShadCR,n=6)
```

These data will be used herein to examine (1) accuracy by comparing the estimated ages from a biologist to the known ages (for simplicity, only the first estimate for a biologist will be used) and (2) precision by comparing between the two estimates by the same biologist and also by comparing among the first estimates of all three biologists. Note that bias is examined as with accuracy, except that two estimated ages are compared rather than one estimated age compared to the true age. Thus, separate examples for bias are not shown here.

# Accuracy/Bias Assessment
## Age-Agreement Table
An age-agreement table is a contingency table that shows how two sets of ages are related. The main diagonal in an age-agreement table represents the individuals for which the two ages are the same. While it ultimately does not make a difference which set of estimated ages form the rows and columns of the age-agreement table, I prefer to have the known or more reliable ages (I call them the "reference ages") form the columns.

An age-agreement table can be formed with `FSA` by using `ageBias()` with the first argument being a formula of the form `nonrefvar~refvar`, with `nonrefvar` and `refvar` generically representing the variables that contain the non-reference and reference ages, respectively. The data.frame that contains these variables is given in `data=`. Labels for the reference and non-reference variables may be given in `ref.lab=` and `nref.lab=`, respectively. The result from `ageBias()` should be assigned to an object for post-processing.

```{r}
ab.tA1 <- ageBias(agerA1~trueAge,data=ShadCR,
                  ref.lab="True Age",nref.lab="Ager A")
```

`ageBias()` computes a number intermediate results and summary statistics that can be specifically extracted with other functions. For example, the age-agreement table is extracted with `summary()` using the object saved from `ageBias()` and `what="table"`.

```{r}
summary(ab.tA1,what="table")
```

Here, for example, it is seen that of the nine true age-3 fish (i.e., the 2nd column) Ager A assigned one fish as age 2, six as age 3, and two as age 4.

## Age-Bias Plot
An *age-bias plot* is used to identify systematic bias between two sets of ages from the same fish. True ages or ages that are thought to be more accurate will be considered as the "reference ages", as mentioned above. In the traditional age-bias plot, the mean age and associated confidence interval computed from the non-reference ages for each reference age is plotted against the corresponding reference age. The plot is augmented with a  1:1 agreement line (i.e., points where the mean non-reference and reference ages are equal). Confidence intervals that do not capture the agreement line indicate a difference (or bias or inaacuracy) between the mean non-reference and reference ages.

The summary statistics required to construct a traditional age-bias plot are computed by `ageBias()` and stored in the `bias` portion of the assigned object.

```{r}
ab.tA1$bias   # ab.tA1 was computed previously (above)
```

For example, the first row above gives the number, minimum, maximum, mean, and standard error of the first estimated ages of Ager A that were paired with a true age of 3. Additionally there is a t-test, adjusted p-value, and a significance statement for testing whether the mean estimated age is different from the true age of 3. Finally, confidence intervals (defaults to 95%) for the mean estimated age at a true age of 3 is given, with a statement about whether or not a confidence interval could be calculated.^[See [the documentation](https://fishr-core-team.github.io/FSA/reference/ageBias.html) for `ageBias()` for the criterion used to decide if the confidence interval can be calculated.]

A typical age-bias plot (@fig-plotAB1) is created below with `ggplot2` from the results in the`bias` object. The specifics of the code below are as follows.

* The base data used in this plot is the `$bias` data.frame discussed above.
* The 45^o^ agreement line (i.e., slope of 1 and intercept of 0) is added with `geom_abline()`, using a dashed `linetype=` and a gray `color=`. This "layer" is first so that it sits behind the other results.
* Error bars are added with `geom_errorbar()`. The `aes()`thetics here map the true age to the `x=` axis and the lower and upper confidence interval values for the mean estimated age at each true age to `ymin=` and `ymax=`. The `color=` of the lines are mapped to the `sig` variable so that points that are significantly different from the 45^o^ agreement line will have a different color (with `scale_color_manual()` described below). Finally, `width=0` assures that the error bars will not have "end caps."
* Points at the mean estimated age (`y=`) for each true age (`x=`) are added with `geom_point()`. Again, the `color=` and `fill=` are mapped to the `sig` variable so that they will appear different depending on whether the points are significantly different from the 45^o^ agreement line or not. Finally, `shape=21` represents a plotted point as an open circle that is outlined with `color=` and filled with `fill=`.
* `scale_fill_manual()` and `scale_color_manual()` are used to set the colors and fills for the levels in the `sig` variable. Note that `guide="none"` is used so that a legend is not constructed for the colors and fills.
* `scale_x_continuous()` and `scale_y_continuous()` are used to set the labels (with `name=`), axis limits, and axis breaks for the x- and y-axes, respectively. The same limits are used for each axis so that the 45^o^ agreement line can be seen for all observed reference ages. The names are drawn from labels that were given in the original call to `ageBias()` and stored in `ab.tA1`.

```{r}
#| label: fig-plotAB1
#| fig-cap: Typical age-bias plot between the first estimates of Ager-A and the true ages.
#| fig-alt: Typical age-bias plot between the first estimates of Ager-A and the true ages.
ggplot(data=ab.tA1$bias) +
  geom_abline(slope=1,intercept=0,linetype="dashed",color="gray") +
  geom_errorbar(aes(x=trueAge,ymin=LCI,ymax=UCI,color=sig),width=0) +
  geom_point(aes(x=trueAge,y=mean,color=sig,fill=sig),shape=21) +
  scale_fill_manual(values=c("FALSE"="black","TRUE"="white"),guide="none") +
  scale_color_manual(values=c("FALSE"="black","TRUE"="red3"),guide="none") +
  scale_x_continuous(name=ab.tA1$ref.lab,breaks=0:10,limits=c(3,8)) +
  scale_y_continuous(name=ab.tA1$nref.lab,breaks=0:10,limits=c(3,8))
```

@fig-plotAB1 suggests that there is no systematic inaccuracies between the first age estimates made by Ager A and the true ages of the Shad as all confidence intervals "capture" the 45^o^ agreement line. The same plot for the first estimates of Ager-B demonstrate a systematic inaccuracy (@fig-plotAB2), noting the color coding for the points and intervals for ages that depict a significant difference.

```{r}
#| label: fig-plotAB2
#| fig-cap: Typical age-bias plot between the first estimates of Ager-B and the true ages.
#| fig-alt: Typical age-bias plot between the first estimates of Ager-B and the true ages.
ab.tB1 <- ageBias(agerB1~trueAge,data=ShadCR,
                  ref.lab="True Age",nref.lab="Ager B")

ggplot(data=ab.tB1$bias) +
  geom_abline(slope=1,intercept=0,linetype="dashed",color="gray") +
  geom_errorbar(aes(x=trueAge,ymin=LCI,ymax=UCI,color=sig),width=0) +
  geom_point(aes(x=trueAge,y=mean,color=sig,fill=sig),shape=21) +
  scale_fill_manual(values=c("FALSE"="black","TRUE"="white"),guide="none") +
  scale_color_manual(values=c("FALSE"="black","TRUE"="red3"),guide="none") +
  scale_x_continuous(name=ab.tB1$ref.lab,breaks=0:10,limits=c(3,8)) +
  scale_y_continuous(name=ab.tB1$nref.lab,breaks=0:10,limits=c(3,8))
```

## Age-Difference Plot
An *age-difference plot* is similar to age-bias plot except that the mean **difference** (and confidence intervals) between the non-reference and reference ages are plotted on the y-axis (@fig-plotAD1). With this the agreement line is a horizontal line at 0. Construction of this plot is similar to above except that the summary results needed for the plot are in `bias.diff` of the object assigned from `ageBias()`.

```{r}
ab.tA1$bias.diff   # ab.tA1 was computed previously (above)
```

In addition, `geom_hline()` is used to add the horizontal agreement line at 0, the breaks for the y-axis need to be adjusted, and the limits for both axes usually need not be specified.

```{r}
#| label: fig-plotAD1
#| fig-cap: Age-difference plot for the first estimates of Ager-A and the true ages.
#| fig-alt: Age-difference plot for the first estimates of Ager-A and the true ages.
ggplot(data=ab.tA1$bias.diff) +
  geom_hline(yintercept=0,linetype="dashed",color="gray") +
  geom_errorbar(aes(x=trueAge,ymin=LCI,ymax=UCI,color=sig),width=0) +
  geom_point(aes(x=trueAge,y=mean,color=sig,fill=sig),shape=21) +
  scale_fill_manual(values=c("FALSE"="black","TRUE"="white"),guide="none") +
  scale_color_manual(values=c("FALSE"="black","TRUE"="red3"),guide="none") +
  scale_x_continuous(name=ab.tA1$ref.lab,breaks=0:10) +
  scale_y_continuous(name=ab.tA1$nref.lab,breaks=-3:2)
```

## Modifications
Both the age-bias and age-difference plots can be modified in a variety of ways. Some options are demonstrated [here](https://fishr-core-team.github.io/fishR/blog/posts/2021-3-15-AgeBiasPlots/#more-examples) and [here](https://fishr-core-team.github.io/fishR/blog/posts/2023_2_13_Quistetal2022_AgeData/). In addition, some authors have suggested using a Bland-Altman plot, which, along with a modification that I prefer, is illustrated [here](https://fishr-core-team.github.io/fishR/blog/posts/2017-4-20_BlandAltman/).

## Bias Metrics
Modern statistical tests for bias focus on whether an age-agreement table is symmetric around the main diagonal. An age-agreement table is a contingency table that shows how two sets of ages are related. The main diagonal in an age-agreement table represents the individuals for which the two ages are the same. If there is no systematic bias between the two ages, then individuals where the paired ages disagree are equally likely to be above or below this main diagonal. Thus, the age-agreement table is *symmetric* around the main diagonal if there is no systematic bias between the two paired ages. [Ogle (2016)](https://derekogle.com/IFAR/) described three statistical tests for assessing the symmetry of an age-agreement table -- *McNemar test* (also known as *maximally pooled*), *Evans-Hoenig test* (also known as *diagonally projected*), and *Bowker test* (also known as *unpooled*).  

Calculating all three tests of symmetry is a two-step process that begins with `ageBias()` as described previously for the age-bias plot. Results from the symmetry tests are displayed by submitting the `ageBias()` object to `summary()` along with `what="symmetry"`.  Alternatively, use `what="McNemar"`, `what="EvansHoenig"`, or `what="Bowker"` to show results from the individual symmetry tests.

```{r}
summary(ab.tA1,what="symmetry")   # ab.tA1 was computed previously (above)
```

The large p-values for all three symmetry tests suggest that the age estimates of Ager A were not systematically biased relative to the true ages (i.e., they were accurate).^[It would be useful to make an age-bias plot here as well.]

To partially illustrate the range of results that may appear in age-agreement tables and tests of symmetry, the first estimated ages from Agers B and C were also compared to the true ages. Ager B appeared to overestimate the age of young fish and underestimate the age of older fish, resulting in a significantly asymmetric age-agreement table, but only as detected by the Bowker test.

```{r}
# note that ab.tB1 was constructed above
summary(ab.tB1,what="table")
summary(ab.tB1,what="symmetry")
```

In contrast, Ager C appeared to underestimate the age of all fish after age 3, resulting in a significantly asymmetric age-agreement table as detected by all three symmetry tests.

```{r}
ab.tC1 <- ageBias(agerC1~trueAge,data=ShadCR,
                  ref.lab="True Age",nref.lab="Ager C")
summary(ab.tC1,what="table")
summary(ab.tC1,what="symmetry")
```

# Precision
## Precision Metrics
There are three commonly used measures of precision among multiple age estimates made on the same fish. *Percent agreement* is the percentage of age estimates for the same fish that either perfectly agree or agree within a certain amount (e.g., within two years of each other). *Average percent error* (APE) is defined as

$$ APE = 100 * \frac{\sum_{j=1}^{n}\sum_{i=1}^{R}\frac{\left|x_{ij}-\bar{x}_{j}\right|}{\bar{x}_{j}}}{nR} $$

where $x_{ij}$ is the $i$th age for the $j$th fish, $\bar{x}_{j}$ is the mean age for the $j$th fish, $R$ is the number of times that each fish was aged (assumed to be the same for all fish), and $n$ is the number of aged fish in the sample. Finally, the *average coefficient of variation* (ACV) is defined as

$$ ACV = 100 * \frac{\sum_{j=1}^{n}\frac{s_{j}}{\bar{x}_{j}}}{n} $$

where $s_{j}$ is the standard deviation of the $R$ age estimates for the $j$th fish.

All three measures of precision are calculated in a two-step process that begins with `agePrecision()` from `FSA`. The first argument to `agePrecision()` is a formula of the form `~var1+var2+...`, where `var1`, `var2`, etc. are the variables that contain the multiple age estimates.^[If only two variables are being considered, then the formula to `agePrecision()` may be the same as that used in `ageBias()`.] The data.frame that contains the variables must be given in `data=`. Again, the result should be assigned to an object for post-processing. For example, precision between the two age estimates made by Ager A is assessed below.

```{r}
ap.A <- agePrecision(~agerA1+agerA2,data=ShadCR)
```

Two types of percentage agreement tables may be extracted by submitting the `agePrecision()` object to `summary()`. The percentage table of raw differences is extracted with `what="difference"`.

```{r}
summary(ap.A,what="difference")
```
```{r}
#| echo: false
sumA1 <- ap.A$rawdiff
sumA1 <- sumA1/sum(sumA1)*100
```

These results show that both estimates by Ager A agreed for `r formatC(sumA1["0"],format="f",digits=1)`\% of the fish and that the second estimate was two years older for `r formatC(sumA1["-2"],format="f",digits=1)`\% of the fish, but two years younger for `r formatC(sumA1["2"],format="f",digits=1)`\% of the fish.

Use `what="absolute difference"` to extract the percentage table of the absolute value of the differences.

```{r}
summary(ap.A,what="absolute difference")
```
```{r}
#| echo: false
sumA2 <- ap.A$absdiff
sumA2 <- sumA2/sum(sumA2)*100
```

These results show that the two age estimates agreed for `r formatC(sumA2["0"],format="f",digits=1)`\% of the fish (as above) and differed by two years (either way) for `r formatC(sumA2["2"],format="f",digits=1)`\% of the fish.

The APE and ACV (along with average standard deviation (ASD) and average absolute deviation (AAD)) are extracted by including `what="precision"` in `summary()`.

```{r}
summary(ap.A,what="precision")
```

The age precision results may be computed for more than two age estimates.  For example, the results below examine precision among the first readings made by all three agers.^[The sample size is much smaller in the example where the three agers are compared because Ager C did not estimate an age for several fish.]

```{r}
ap.ABC <- agePrecision(~agerA1+agerB1+agerC1,data=ShadCR)
summary(ap.ABC,what="difference")
summary(ap.ABC,what="precision")
```

# Other Considerations
## Built-In Age-Bias Plot
A quick-and-dirty, not publication quality, traditional age-bias plot may be constructed from `ageBias()` object with `plotAB()`.

```{r}
plotAB(ab.tA1)
```


## Individual Fish
The objects returned from `ageBias()` and `agePrecision()` contain intermediate calculations for individual fish that can be used for further analysis. For example, `data` in the `ageBias()` object contains the original two age estimates along with the difference and the mean of those two estimates. Some authors have used these data to create mixed models to assess components that might contribute to ageing error ([here is one example](https://meridian.allenpress.com/jfwm/article/15/1/203/501110/Use-of-Fin-Rays-and-Fin-Spines-in-Nonlethal-Age)).

```{r}
peek(ab.tA1$data,n=6)
```

`detail` in the `agePrecision()` object contains a variety of information related to the estimates for the individual fish, including the individual level CV and PE.^[See [documentation](https://fishr-core-team.github.io/FSA/reference/agePrecision.html) for `agePrecision()` for more details.] One example of individual-based analyses that may derive from these data is [shown here](https://fishr-core-team.github.io/fishR/blog/posts/2015-9-5_Age_Comparison_Results/).

```{r}
peek(ap.A$detail,n=6)
```
